Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
more comments and some changes

@@

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one which resulted in 27,888 genes remaining.

To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means. We estimated sample-specific normalization factors to account for sequencing depth. Then, we estimated common a common dispersion for each gene, but shared across the samples. Finally, we obtained independent estimates for the gene-variety specific location parameters. The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study, i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.

@@

Existing RNAseq methods are not designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide a statistic representing the plausibility of heterosis, i.e. higher values indicating higher plausibility.

The approach used by {\tt edgeR} naturally provides testing of hypotheses for differential expression amongst two varieties \citep{robinson2007moderated, robinson2010edgeR}. If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis. We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2. But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e. mid-parent heterosis, and therefore not be exhibiting extreme heterosis. We construct a statistic that is 0 if the estimate of mean expression 
for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid. 
% NEED MLE

>>>>

% WML: Please clarify that {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} are R packages. You may even want to cite R somewhere.

@@

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, which resulted in 27,888 genes remaining.

% WML: Sentences 2-4 in the paragraph below are redundant with sentence 1. I commented out the paragraph below and inserted a modified one afterwards
%To obtain gene-variety-specific location parameters and gene-specific dispersion parameters, we used {\tt edgeR} in order to maintain existing covariance structure amongst the gene-variety means. We estimated sample-specific normalization factors to account for sequencing depth. Then, we estimated common a common dispersion for each gene, but shared across the samples. Finally, we obtained independent estimates for the gene-variety specific location parameters. The location parameters, dispersion parameters, and normalization factors were then treated as the truth for the simulation study, i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH.

To begin, we applied the {\tt edgeR} package from Bioconductor (citation needed) to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated all parameters as truth for the simulation study:  i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide statistics to measure the strength of heterosis of each gene.

% WML: "... provide a statistic representing the plausibility of heterosis, i.e. higher values indicating higher plausibility." => "...provide statistics to measure the strength of heterosis of each gene."

% WML: change: "edgeR was originally designed to..."

{\tt edgeR} was originally designed to test for differential expression between two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a statistic for each gene. If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis. We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2. But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e. mid-parent heterosis, and therefore not be exhibiting extreme heterosis. We construct a statistic that is 0 if the estimate of mean expression 
for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.

% NEED MLE
